Hydras and Cie \\ Variations on Kirby \& Paris' hydra battles and other entertaining math in Coq

### 2021-09-10 doc/JFLA2022/paper.tex
draft paper

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

@@

@@

\section{New contents}
Length of usually considered battles.
They correspond to the notions of canonical sequences and large sets in \cite{KS81}.

\begin{todo}
  Relation with Russel's library (maintained in Coq-community).
\end{todo}

@@

\label{Ex42-schutte}
\fbox{\begin{minipage}[h]{1.0\linewidth}
  Let us prove again the equality $\omega+42+\omega^2= \omega^2$.Let us recall that $\omega^2$ is an abbreviation of $\phi_0(2)$,
\emph{i.e} the third  additive principal ordinal.

@@

\section{Perspectives}
The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementaion in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic) definitions of
\texttt{T1} and \texttt{T2}, originated in~\cite{CantorContrib}.
The types --- defined in different modules --- cannot be considered as equal. Even more, some operations like addition, multiplication, etc. have now different definitions.

>>>>

% Some of our commands for this guide.
%
\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}
\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}

@@

Our project is to illustrate  this work by a library written in \emph{Coq} and an electronic book containing explanations of the mathematical content and its possible mechanized formalisations.

@@

\begin{todo}
  Comment the multiple views of a same mathematical object:
  \begin{itemize}
  \item Ordinals notations: for $\omega$, $\omega^\omega$,
    $\epsilon_0$, $\Gamma_0$. Proofs by computation whenever possible (see Fig.~\ref{Ex42-E0}.
  \item (Axiomatic) model by Schütte. Less computations.
    application of previously proven lemmas (see Fig~\ref{fig:Ex42b}).
  \item Bridge lemmas (relative correctness of different representations).
    \item The user/reader may compare  different proofs of the same result (For instance the equality $\omega+42+\omega^2= \omega^2$).
  \end{itemize}
\end{todo}

\section{New contents}
We are interested in a usual variant of hydra-battles, where the hydra grows $n$ copies at step $n$ of the game (see the examples presented in ~\cite{KP82, BauerHydra}).
In terms of ordinals, this variant is described as a (finite) sequence $s_n\,(n\geq i)$ of ordinals strictly below $\epsilon_0$.
\begin{itemize}
\item If $s_n=0$, then the sequence is empty,
\item if $s_n=\beta+1$, then $s_{n+1}=\beta$,
\item Otherwise, $s_n$ is a \emph{limit ordinal} $\lambda$; then $s_{n+1}$ is $\canonseq{\lambda}{n+1}$, \emph{i.e.} the
    $n+1$-th element of the \emph{canonical sequence} associated with $\lambda$~\cite{KS81}.
\end{itemize}

The length of such a sequence is $l-i$, where $s_l=0$. It depends both on the initial index $i$ and the initial term $\alpha=s_i$. Adapting to \emph{Coq} a few lemmas from~\cite{KS81}, we prove that this length is minored by a fast growing function.
More precisely, the length  of the sequence associated with
$\omega^\alpha$ and initial index $i\geq 2$ is greater or equal than
$F_{\omega^\alpha}(i-1)-(i-1)$, where $F_\alpha$ is a variant of the Wainer hierarchy of rapidly growing functions, defined and studied in~\cite{KS81}.
From this minoration, we infer that the function which associates to $i$ the length of the considered sequence (for a given $\alpha$) is not primitive recursive in general (our counter-example is for $\alpha=\omega^\omega$).

It is important to notice that, for proving this last result, we decided to host and maintain within \emph{Coq-Community} the formalization of primitive recursive function, part of Russel O'Connor work on G\"{o}del's first incompleteness theorem~\cite{OConnor05}.

Defining and studying families of rapidly growing arithmetic  functions indexed by ordinals below $\epsilon_0$ has been made possible by the plug-in \emph{coq-equations}~\cite{sozeau:hal-01671777}.

\section{The Coq-community ecosystem}

@@

\label{Ex42-schutte}
\fbox{\begin{minipage}[h]{1.0\linewidth}
  Let us prove again the equality $\omega+42+\omega^2= \omega^2$. Let us recall that $\omega^2$ is an abbreviation of $\phi_0(2)$,
\emph{i.e} the third  additive principal ordinal.

@@

\section{Conclusion and perspectives}

\begin{todo}
  \begin{itemize}
  \item  Soundness (correction, maintenance) of the Alectryon generated doc.
  \item Communications between co-maintained libraries
          (Ackermann, Gaia, etc.)
  \item Work destined both to Coq-users and people interested in discrete math ?
  \end{itemize}
\end{todo}

\subsection{Perspectives}

The Gaia project~\cite{Gaia}, also maintained on \textit{Coq-community} contains a development by José Grimm~\cite{grimm:hal-00911710}. This library -- dedicated  to the implementaion in \textit{Coq (SSreflect)} of books from  Bourbaki's Elements of Mathematics --- contains a generalization of two types of ordinals (\texttt{T1} for the ordinal $\epsilon_0$ and
\texttt{T2} for $\Gamma_0$) to a larger ordinal defined by Ackermann, and relates these data structures with Bourbaki's set theory.
It is interesting to note that both libraries \textit{Gaia} and \textit{hydra-battles} contain their own (isomorphic) definitions of
\texttt{T1} and \texttt{T2}, originated in~\cite{CantorContrib}.
The types --- defined in different modules --- cannot be considered as equal. Even more, some operations like addition, multiplication, etc. have now different definitions.
